Packet-by-Packet Hardening of Intrusion Detection Systems against Smart Adversaries

### 2019-10-25 ex3/paper/paper.tex
Paper draft

\newacronym{ml}{ML}{Machine Learning}
\newacronym{dl}{DL}{Deep Learning}
\newacronym{aml}{AML}{Adversarial Machine Learning}
\newacronym{ids}{IDS}{Intrusion Detection System}

@@

We can confirm that adversarial samples can be found even when considering the tight constraints mentioned before. Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}. We develop a simple methods for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process. The resulting classifier has slightly reduces accuracy but is still useful to be deployed in a real scenario. Finally we develop a custom adversarial training procedure which especially focuses on easy of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy we can reduce attack surface and harden the resulting \gls{ids} while keeping similar classification performance like for the original \gls{ml} model.

@@

@@

As we saw in the previous section in \autoref{fig:plot_features}, some attack types do have a characteristic pattern of interarrival times and packet lengths. Furthermore, \autoref{fig:pred_plots2} shows after the first couple of packets, the classifier becomes so sure in its decision, that a change in features doesn't seem to have any after the first packets. The hypothesis that follows is that by manipulating the interarrival times and packet lengths of the first packets, the classifier might be fooled. However, our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot manipulated freely: Besides the constraints that packet size and interarrival time cannot become negative there are also more problem-specific constraints:
\begin{itemize}
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item Interarrival times must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of interarrival times is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

\subsection{Carlini-Wagner}
In particular, we implemented the method by \citeauthor{carlini2017towards}~\cite{carlini2017towards}, which performs gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + \epsilon  \max(Z(\tilde X), 0).
\end{equation}
Here, $d(\cdot)$ is a distance metric, $\epsilon \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.

We used $\ell_1$ norm as distance metric and deployed projected gradient descent (PGD) for meeting the real-world constraints discussed above.

@@

\subsection{$\ell_\infty$-bounded Projected Gradient Descent}
\cite{madry2017towards} use a method for generating adversarial samples which uses PGD  to minimize the network's negative loss function while constraining the achieved $\ell_\infty$ distance from the original samples.

\subsection{Fast Gradient Sign Method}
Finally, we also tested the Fast Gradient Sign Method (FGSM), which was initially proposed in \cite{goodfellow2014explaining} and perhaps is the most frequently used method for generating adversarial sampling. FGSM can be considered a single pass of PGD on the loss function with an equality constraint on the $\ell_\infty$ distance, i.e. the adversarial sample is found as
\begin{equation}
\tilde X = X + \epsilon \text{sgn}( \nabla_X L(X)),
\end{equation}
where $L(X)$ denotes the network's loss function and $\epsilon$ denotes the achieved $\ell_\infty$ distance.

\subsection{Evaluation}
\begin{figure*}
\includegraphics[width=0.45\textwidth]{adv_plots/cwl1.pdf}
\includegraphics[width=0.45\textwidth]{adv_plots/cwl2.pdf}
\caption{Performance of the Carlini-Wagner method with different $\epsilon$ values for $\ell_1$ and $\ell_2$ distances.}
\label{fig:adv_cw}
\end{figure*}
\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as normal traffic with respect to the final packet along with the resulting distance from the original sample. Evidently, the larger the $\epsilon$ parameter, the better the attack works, but also the distances become larger.

@@

\begin{table*}
\caption{Attack success ratios per attack family. Depicted are flow detection accuracies for adversarial flows (left numbers) and flow detection accuracies for unmodified flows (right numbers).}
\label{tab:adv_per_family}
\centering
\begin{tabular}{lllllllll}
\toprule
Attack type & FGSM & Carlini-Wagner, $\ell_1$ & Carlini-Wagner, $\ell_2$ & $\ell_\infty$-bounded PGD \\
\midrule
Botnet:ARES	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Brute Force:FTP-Patator	&	0.2800 / 0.5800	&	0.1600 / 0.5800	&	0.1600 / 0.5800	&	0.2800 / 0.5800	\\
Brute Force:SSH-Patator	&	0.8300 / 1.0000	&	0.0000 / 1.0000	&	0.0000 / 1.0000	&	0.8900 / 1.0000	\\
DDoS:LOIT	&	0.8300 / 0.8600	&	0.6400 / 0.8600	&	0.5600 / 0.8600	&	0.8200 / 0.8600	\\
DoS / DDoS:DoS GoldenEye	&	0.8900 / 1.0000	&	0.5000 / 1.0000	&	0.4200 / 1.0000	&	0.9200 / 1.0000	\\
DoS / DDoS:DoS Hulk	&	0.6700 / 1.0000	&	0.5400 / 1.0000	&	0.4700 / 1.0000	&	1.0000 / 1.0000	\\
DoS / DDoS:DoS Slowhttptest	&	0.8700 / 1.0000	&	0.4100 / 1.0000	&	0.1900 / 1.0000	&	0.7400 / 1.0000	\\
DoS / DDoS:DoS slowloris	&	0.4400 / 0.9700	&	0.3100 / 0.9700	&	0.1300 / 0.9700	&	0.9200 / 0.9700	\\
DoS / DDoS:Heartbleed	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Infiltration:Dropbox download 	&	0.6100 / 0.9600	&	0.4300 / 0.9600	&	0.3800 / 0.9600	&	0.7400 / 0.9600	\\
PortScan:PortScan - Firewall off	&	0.6800 / 0.9100	&	0.3500 / 0.9100	&	0.3500 / 0.9100	&	0.7200 / 0.9100	\\
PortScan:PortScan - Firewall on	&	0.7400 / 0.8700	&	0.6100 / 0.8700	&	0.5300 / 0.8700	&	0.7300 / 0.8700	\\
Web Attack:Sql Injection	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	\\
Web Attack:XSS	&	0.6700 / 0.9500	&	0.7200 / 0.9500	&	0.6200 / 0.9500	&	0.6700 / 0.9500	\\
\bottomrule

@@

\autoref{fig:fgsm} depicts the performance for FGSM and $\ell_\infty$-bounded PGD. As expected, while FGSM is the fastest of all algorithms, it delivers the worst performance.

Of the investigated algorithms, thus the Carlini-Wagner method with $\ell_2$ distance delivered the best performance.

\autoref{tab:adv_per_family} depicts how successful the attack is for individual attack families. For this table we used a $\ell_\infty$ bound of 0.3 for FGSM and $\ell_\infty$-bounded PGD and a tradeoff of 0.5 for Carlini-Wagner. The table shows significant for detecting different attack families in the first place. However, also for generating adversarial flows some attack families are more susceptible than others.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{adv_plots/fgsm.pdf}
\includegraphics[width=0.45\textwidth]{adv_plots/l_inf_pgd.pdf}
\caption{Performance of FSGM and $\ell_\infty$-bounded PGD for different bounds.}
\label{fig:fgsm}
\end{figure*}

@@

To this end, we generated adversarial flows with the Carlini-Wagner method with $\ell_2$ distance and a tradeoff of 0.5. \autoref{adv_plots} depicts how the method manipulates flows for some exemplary attack categories. Depicted is the mean of both unmodified flows and adversarial flows  of samples of individual attack families considering both interarrival time and packet length.

@@

\begin{figure*}[p]
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/1.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/2.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/3.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/4.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/5.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/6.pdf}
\caption{Successful manipulations performed by the Carlini-Wagner method.}
\label{adv_plots}
\end{figure*}

@@

In addition to omitting manipulable features we tried to make our network more robust against adversarial samples by augmenting the training set by adversarial flows generated using the Carlini-Wagner method, labeled as additional attacks. However, against our expectations, this approach did not lead to satisfactory results in the sense that the network would learn to classify adversarial samples as attacks. In fact, classification accuracy stayed about constant during the training of the network.

>>>>

\newacronym{ml}{ML}{Machine Learning}
\newacronym{dl}{DL}{Deep Learning}
\newacronym{aml}{AML}{Adversarial Machine Learning}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{rnn}{RNN}{Recurrent Neural Network}
\newacronym{fgsm}{FGSM}{Fast Gradient Sign Method}
\newacronym{cw}{CW}{Carlini-Wagner method}
\newacronym{pgd}{PGD}{Projected Gradient Descent}

@@

We can confirm that adversarial samples can be found even when considering the tight constraints mentioned before. Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}. We develop a simple methods for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process. The resulting classifier is slightly less accurate but is still useful to be deployed in a real scenario. Finally we develop a custom adversarial training procedure which especially focuses on easy of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy we can reduce attack surface and harden the resulting \gls{ids} while keeping similar classification performance like for the original \gls{ml} model.

@@

\cite{szegedy_intriguing_2014} show that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly. \cite{goodfellow_explaining_2015} develop the \gls{fgsm}, which makes easy generation of adversarial samples possible. 
\cite{papernot_crafting_2016} explore the use of \gls{aml} for \glspl{rnn}

@@

Since most features such as the TCP flags cannot be easily be manipulated because it might violate the TCP and thus cause communication to fails, we experiment whether by manipulating the interarrival times and packet lengths of the first packets, the classifier might be fooled. However, our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot manipulated freely: Besides the constraints that packet size and interarrival time cannot become negative there are also more problem-specific constraints:
\begin{itemize}
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item Interarrival times must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of interarrival times is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

\subsection{Carlini-Wagner}
In particular, we implemented \gls{cw} \cite{carlini_towards_2017}, which performs gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + \epsilon  \max(Z(\tilde X), 0).
\end{equation}
Here, $d(\cdot)$ is a distance metric, $\epsilon \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.

We used $\ell_1$ norm as distance metric and \gls{pgd} for meeting the real-world constraints discussed above.

@@

\subsection{$\ell_\infty$-bounded Projected Gradient Descent}
\cite{madry_towards_2018} use a method for generating adversarial samples which uses \gls{pgd}  to minimize the network's negative loss function while constraining the achieved $\ell_\infty$ distance from the original samples.

\subsection{Fast Gradient Sign Method}
Finally, we also tested the Fast Gradient Sign Method (FGSM), which was initially proposed in \cite{goodfellow_explaining_2015} and perhaps is the most frequently used method for generating adversarial sampling. FGSM can be considered a single pass of \gls{pgd} on the loss function with an equality constraint on the $\ell_\infty$ distance, i.e. the adversarial sample is found as
\begin{equation}
\tilde X = X + \epsilon \text{sgn}( \nabla_X L(X)),
\end{equation}
where $L(X)$ denotes the network's loss function and $\epsilon$ denotes the achieved $\ell_\infty$ distance.

\subsection{Evaluation}
\begin{figure*}
\includegraphics[width=0.45\textwidth]{adv_plots/cwl1.pdf}
\includegraphics[width=0.45\textwidth]{adv_plots/cwl2.pdf}
\caption{Performance of the \gls{cw} with different $\epsilon$ values for $\ell_1$ and $\ell_2$ distances.}
\label{fig:adv_cw}
\end{figure*}
\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as normal traffic with respect to the final packet along with the resulting distance from the original sample. Evidently, the larger the $\epsilon$ parameter, the better the attack works, but also the distances become larger.

@@

\begin{table*}
\caption{Attack success ratios per attack family. Depicted are flow detection accuracies for adversarial flows (left numbers) and flow detection accuracies for unmodified flows (right numbers).}
\label{tab:adv_per_family}
\centering
\begin{tabular}{lllllllll}
\toprule
Attack type & FGSM & Carlini-Wagner, $\ell_1$ & Carlini-Wagner, $\ell_2$ & $\ell_\infty$-bounded \gls{pgd} \\
\midrule
Botnet:ARES	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Brute Force:FTP-Patator	&	0.2800 / 0.5800	&	0.1600 / 0.5800	&	0.1600 / 0.5800	&	0.2800 / 0.5800	\\
Brute Force:SSH-Patator	&	0.8300 / 1.0000	&	0.0000 / 1.0000	&	0.0000 / 1.0000	&	0.8900 / 1.0000	\\
DDoS:LOIT	&	0.8300 / 0.8600	&	0.6400 / 0.8600	&	0.5600 / 0.8600	&	0.8200 / 0.8600	\\
DoS / DDoS:DoS GoldenEye	&	0.8900 / 1.0000	&	0.5000 / 1.0000	&	0.4200 / 1.0000	&	0.9200 / 1.0000	\\
DoS / DDoS:DoS Hulk	&	0.6700 / 1.0000	&	0.5400 / 1.0000	&	0.4700 / 1.0000	&	1.0000 / 1.0000	\\
DoS / DDoS:DoS Slowhttptest	&	0.8700 / 1.0000	&	0.4100 / 1.0000	&	0.1900 / 1.0000	&	0.7400 / 1.0000	\\
DoS / DDoS:DoS slowloris	&	0.4400 / 0.9700	&	0.3100 / 0.9700	&	0.1300 / 0.9700	&	0.9200 / 0.9700	\\
DoS / DDoS:Heartbleed	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Infiltration:Dropbox download 	&	0.6100 / 0.9600	&	0.4300 / 0.9600	&	0.3800 / 0.9600	&	0.7400 / 0.9600	\\
PortScan:PortScan - Firewall off	&	0.6800 / 0.9100	&	0.3500 / 0.9100	&	0.3500 / 0.9100	&	0.7200 / 0.9100	\\
PortScan:PortScan - Firewall on	&	0.7400 / 0.8700	&	0.6100 / 0.8700	&	0.5300 / 0.8700	&	0.7300 / 0.8700	\\
Web Attack:Sql Injection	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	\\
Web Attack:XSS	&	0.6700 / 0.9500	&	0.7200 / 0.9500	&	0.6200 / 0.9500	&	0.6700 / 0.9500	\\
\bottomrule

@@

\autoref{fig:fgsm} depicts the performance for FGSM and $\ell_\infty$-bounded \gls{pgd}. As expected, while FGSM is the fastest of all algorithms, it delivers the worst performance.

Of the investigated algorithms, thus the \gls{cw} with $\ell_2$ distance delivered the best performance.

\autoref{tab:adv_per_family} depicts how successful the attack is for individual attack families. For this table we used a $\ell_\infty$ bound of 0.3 for FGSM and $\ell_\infty$-bounded \gls{pgd} and a tradeoff of 0.5 for Carlini-Wagner. The table shows significant for detecting different attack families in the first place. However, also for generating adversarial flows some attack families are more susceptible than others.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{adv_plots/fgsm.pdf}
\includegraphics[width=0.45\textwidth]{adv_plots/l_inf_pgd.pdf}
\caption{Performance of FSGM and $\ell_\infty$-bounded \gls{pgd} for different bounds.}
\label{fig:fgsm}
\end{figure*}

@@

To this end, we generated adversarial flows with the \gls{cw} with $\ell_2$ distance and a tradeoff of 0.5. \autoref{adv_plots} depicts how the method manipulates flows for some exemplary attack categories. Depicted is the mean of both unmodified flows and adversarial flows  of samples of individual attack families considering both interarrival time and packet length.

@@

\begin{figure*}[p]
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/1.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/2.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/3.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/4.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/5.pdf}
\includegraphics[width=0.48\textwidth]{../plots/plot_adv/6.pdf}
\caption{Successful manipulations performed by the \gls{cw}.}
\label{adv_plots}
\end{figure*}

@@

In addition to omitting manipulable features we tried to make our network more robust against adversarial samples by augmenting the training set by adversarial flows generated using the \gls{cw}, labeled as additional attacks. However, against our expectations, this approach did not lead to satisfactory results in the sense that the network would learn to classify adversarial samples as attacks. In fact, classification accuracy stayed about constant during the training of the network.
